# Advanced treatment assignments

So far, we have focused on estimation for a fixed treatment assignment: assign to treatment 1 with probability 1.

We might also want to know about the gap-closing estimand

- if we assigned people to treatment stochastically
- if each person's assignment were individualized

## Stochastic treatments

We may want to study a counterfactual where treatment is assigned with some probability between 0 and 1. The `counterfactual_assignments` argument can handle this possibility.

For example, consider the gap-closing estimand if assigned to treatment 1 with each probability .75.
```{r}
estimate_stochastic <- gapclosing(
  data = simulated_data,
  counterfactual_assignments = .75,
  outcome_formula = formula(outcome ~ confounder + category*treatment),
  treatment_formula = formula(treatment ~ confounder + category),
  category_name = "category"
)
```
The disparity between categories A and B under that stochastic intervention (0.75 probability of treatment = 1) is estimated to be `r round((estimate_stochastic$counterfactual_disparities %>% filter(category == "B - A"))$estimate,2)`, whereas under the previous deterministic intervention to assign treatment to the value 1 the disparity would be `r round((estimate$counterfactual_disparities %>% filter(category == "B - A"))$estimate,2)`. This illustrates an important point: the gap-closing estimand can be different depending on the counterfactual assignment rule, as the figure below shows for counterfactuals in which treatment is assigned with probabilities ranging from 0 to 1.
```{r, echo = FALSE, fig.caption = "A conceptual figure for which the package does not provide a general-purpose function."}
counterfactual_assignments_values <- seq(0,1,.25)
many_stochastic_estimates <- rep(NA, length(counterfactual_assignments_values))
for (i in 1:length(counterfactual_assignments_values)) {
  estimate_case <- gapclosing(
    data = simulated_data,
    counterfactual_assignments = counterfactual_assignments_values[i],
    outcome_formula = formula(outcome ~ confounder + category*treatment),
    treatment_formula = formula(treatment ~ confounder + category),
    category_name = "category"
  )
  estimate_we_want <- estimate_case$change_disparities %>%
    filter(change_type == "proportional") %>%
    filter(category == "B - A")
  many_stochastic_estimates[i] <- estimate_we_want$estimate
}
data.frame(counterfactual_assignments = counterfactual_assignments_values,
           estimate = many_stochastic_estimates) %>%
  ggplot(aes(x = counterfactual_assignments,
             y = estimate)) +
  geom_hline(yintercept = c(0,1),
             color = "gray", linetype = "dashed") +
  geom_line(color = "gray") +
  geom_point(aes(color = factor(counterfactual_assignments == .75))) +
  xlab("Counterfactual Treatment Assignment Probability") +
  scale_y_continuous(name = "Percent of B - A Gap Closed",
                     labels = function(x) paste0(round(100*x),"%")) +
  annotate(geom = "text",
           x = 1, y = c(0,1),
           label = c("No change to disparity",
                     "Disparity completely eliminated"),
           color = "darkgray", size = 3, hjust = 1, vjust = 1.5) +
  annotate(geom = "text",
           x = .1, y = .3, 
           label = "Disparities eliminated under various\nstochastic\ninterventions",
           color = "gray", size = 3, hjust = 0, vjust = 1) +
  annotate(geom = "text",
           x = .75, y = .3,
           label = "Gap-Closing Estimand\nCalculated Above",
           color = "blue", size = 3, vjust = -.5) +
  scale_color_manual(values = c("gray","blue")) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = "none")
```

## Individualized treatments

Treatment may also be different (and possibly stochastic) for each unit in the counterfactual world of interest.

For example, suppose we assign those in Category A to treatment 1 with probability .5, those in Category B to treatment with probability .4, and those in Category C to treatment with probability .3. In this case, `counterfactual_assignments` will be set to a vector of length `nrow(data)`.
```{r}
our_assignments <- case_when(simulated_data$category == "A" ~ .5,
                             simulated_data$category == "B" ~ .4, 
                             simulated_data$category == "C" ~ .3)
estimate_stochastic <- gapclosing(
  data = simulated_data,
  counterfactual_assignments = our_assignments,
  outcome_formula = formula(outcome ~ confounder + category*treatment),
  treatment_formula = formula(treatment ~ confounder + category),
  category_name = "category"
)
```
That intervention would close the B - A gap by `r paste0(round(100 * (estimate_stochastic$change_disparities %>% filter(category == "B - A" & change_type == "proportional"))$estimate),"%")`.